\chapter{Overview}
%\section{What QMC calculates}
%\section{What QMCPACK calculates}
QMCPACK is an object-oriented QMC code which implements zero-temperature QMC algorithms for three dimensional, off-lattice systems and is parallelized with both MPI and OpenMP.  Algorithms currently implemented in QMCPACK are Variational Monte Carlo (VMC)~\cite{PhysRev.138.A442,PhysRevB.16.3081}, Diffusion Monte Carlo (DMC)~\cite{Ceperley1979,Schmidt1984}, and Reptation Monte Carlo (RMC)~\cite{PhysRevLett.82.4745}.  Prior to reading this documentation, the user is assumed to have basic understanding of the Monte Carlo algorithm to be used.  For beginners, a good place to start would be Refs.~\cite{Nightingale1998,Thijssen2007} and references therein.
 
%\subsection{Units}
All numerical values processed in QMCPACK are in atomic units, ie. bohrs, elementary charges, electron masses, and hartrees.  Consult \url{http://en.wikipedia.org/wiki/Atomic_units} for conversion factors with other systems of units.  It is part of the future plan to incorporate a few other units popular for quantum many-body problems and also systems of dimensions lower than 3.

QMCPACK is developed by the Electronic Structure Group at the University of Illinois,\footnote{Current contributors are \href{mailto:jnkim@ncsa.illinois.edu}{Jeongnim Kim}, \href{mailto:esler@illinois.edu}{Ken Esler}, \href{mailto:mcminis2@illinois.edu}{Jeremy McMinis}, \href{mailto:jkrogel2@illinois.edu}{Jaron Krogel}, \href{mailto:bclark@princeton.edu}{Bryan Clark}, \href{mailto:johngergely@gmail.com}{John Gergely}, \href{mailto:Miguel.Morales@rice.edu}{Miguel Morales} and \href{mailto:dcyang2@illinois.edu}{ChangMo Yang}.  Different parts of the code are written by different people, whose names are usually signed at the top of each file.  Be sure contact the correct author of the file in case of any bugs, or any other compiling issues.} funded by the U.S. National Science Foundation and Department of Energy. The software is released under the \href{http://www.opensource.org/licenses/UoI-NCSA.php}{UIUC/NCSA Open Source License} (OSI approved). 
%\section{QMC algorithms implemented}
%\subsection{Variational Monte Carlo}
%\subsection{Diffusion Monte Carlo}
%\subsection{Reptation Monte Carlo}
%\section{Sources of error}
%Brief Overview of QMC
%  What is calculable w/ QMC?
%  Why use QMC over other methods?
%  Modern methods of continuum QMC
%    Overview of VMC Method
%    Overview of DMC Method
%    Overview of RMC Method
%    Sources of Error in QMC Methods
%QMCPACK: an efficient implementation of QMC Methods
%  Quantities obtainable via QMCPACK
%  Scalability of QMCpack
%  Methods implemented in QMCPACK
