\chapter{Getting Started}
%developer contact info here (Jeongnim? or all of us?)
\section{Obtaining QMCPACK}
QMCPACK is written in C++ and is currently under the New BSD License.  Due to its incomplete compliance with the current ISO C++ standard, it is not released in cross-compilable packages.  It is instead obtained from its source code repository hosted by Google Code.  As described in project website at http://code.google.com/p/qmcpack/, the latest revision of the code can be checked out anonymously using Subversion, with the command
\begin{term}
svn checkout http://qmcpack.googlecode.com/svn/trunk/ qmcpack-read-only
\end{term}
where the local directory name \icode{qmcpack-read-only} can be changed as needed.  The repository forbids anonymous users (ie. nonmembers of the project) from committing changes to the code.  The above command will create a directory \icode{qmcpack-read-only} (topdir) which contains
\begin{verbatim} 
qmcpack-read-only/
  CMakeLists.txt
  src/           : directory for the source
    CMakeLists.txt
    dir1/
      CMakeLists.txt
    dir2/
      CMakeLists.txt
    ....
  CMake/        : directory with cmake files for external package and compiler
  config/       : directory with toolchain files for tested systems
  docs/         : directory with Doxygen files
  utils/        : directory with utility scripts
  build/        : empty directory for a build 
\end{verbatim}

%can pretty much copy the svn instructions from google code
\section{Building QMCPACK}
The aim of this section is to provide a guideline for building QMCPACK in the most generic way.
A compromise must be made between overly many case-by-case examples and a simplified procedure.

%Theoretically speaking, 
QMCPACK can be used on any *nix machine with C++ compilers that are reasonably recent.  Some examples to date are g++ $\geq 4.2$ (GNU) and icc $\geq 10.1$ (Intel).  There are other required packages which the compilation process will point out if missing.  Listed below are those required specifically by QMCPACK, with the oldest tested versions.
\begin{itemize*}
\item{} CMake $\geq 2.8$
\item{} mpich, mvapich or openmpi
\item{} blas or lapack library (provided as mkl in Intel compilers)
\item{} boost library
\item{} fftw library $\geq 3$
\item{} xml2 library
\item{} hdf5 library $\geq 1.8$
\item{} pygtkglext: OpenGl extensions for Python GTK bindings for the GUI interface.
% This package has some dependencies.
\end{itemize*}
For library packages, their full development versions (with suffixes -dev or -devel) are needed.  No additional environment variables need to be set if the above packages are installed from precompiled binaries.\footnote{Typical examples are rpm or deb packages.}  If instead the libraries are locally built, their installation directories must be explicitly set as environment variables, eg. \icode{LIBXML2\_HOME=/usr/local}.  See the installation notes in each package for details.  On many HPC centers, these packages are managed by utilities like module and softenv.  Again, see the documentation on each site for details.

% recent [meaning $> 4.2$ at least] GNU
%Compiling the QMCPACK source requires the CMake package.
%For cmake documentation and guides, consult cmake wiki.

Assuming that all of the above packages are properly set up, building QMCPACK proceeds as follows.  Change to qmcpack the top directory (name can be different).
\begin{term}
cd qmcpack-read-only
\end{term}
We recommend out-of-source compilation by creating a directory for the libraries and binaries that is separate from the source directory.  If not in \iterm{qmcpack-read-only} already, create the build directory (eg. \iterm{build}) then change to the directory.\footnote{Like the top directory, the build directory name is completely arbitrary and it can be created at any location, even outside \iterm{qmcpack-read-only}.  This is useful if you need more than one build, for testing purposes.}
\begin{term}
mkdir build
cd build
\end{term}
In the build directory, run cmake to create Makefiles, then build the executable using make.
%(Don't forget ..)
\begin{term}
cmake ..
make
\end{term}
If everything goes well, then you should see \iterm{qmcpack-read-only/build/bin/qmcapp}.
%The procedure above, creating build directory and running camke in a new directory, is an example. We can further separate the source (development) and build. Let's assume that the QMCPACK topdir is /home/foo/src/qmcpack. Then, one can build multiple executables in different locations by creating new directories and build QMCPACK in each directory.
%\begin{verbatim}
%/home/foo/build/gcc-real
%/home/foo/build/gcc-complex
%/home/foo/build/mpi-real
%....
%\end{verbatim}
%
%In each directory, e.g., /home/foo/build/gcc-real (after setting the environments properly), execute
%\begin{verbatim}
%$cmake /home/foo/src/qmcpack
%$make 
%\end{verbatim}
So far, there is no need to change sources or cmake files.  \icode{cmake ..} in the above procedure uses \iterm{..} because the source tree resides in the parent directory.  If something did not work, simply remove the directory (eg. \icode{rm -rf build}) and start again. 

Additional configurations must be considered if you need to deviate from the default settings of parallel computing (MPI) and multithreading (OpenMP).  See \S{}\ref{ss:compset} for details.

%discuss toolchains here
\subsection{Compiler settings for MPI and OpenMP} \label{ss:compset}
%To run QMCPACK under MPI, 
\paragraph{MPI}
Running QMCPACK under MPI requires it to be built with a modified compiler.  This usually involves installing an additional \emph{compiler wrapper} package which turns the already working compiler into an MPI-capable one, without having to rebuild the compiler itself.\footnote{MPI implementations of this type are known as \emph{portable implementations}.}  Examples of MPI implementations known to work with QMCPACK are OpenMPI, MPICH2, and MVAPICH.
%icc vs. g++

Some environment variables must be set in order to use MPI.  By default, the variable CXX is set to the serial compiler (either g++ or icc).  Change this to one of the following:
%MPI is automatically enabled if
\begin{itemize*}
  \item{} mpicxx, mpic++, cmpic++ (tungsten at NCSA)
  \item{} mpCC/mpCC\_r on AIX 
\end{itemize*}
Next, add the directory containing mpi.h to the include file search path.  If the compiler wrappers were installed from binary packages or were locally built with default options, they should already be in one of the directories in the standard search path, eg. /usr/include or /usr/local/include.  If you want to keep MPI enabled for other projects but disable it for QMCPACK only, do one of the following:
%s, eg., /usr/include or /usr/local/include
%  \begin{itemize*}
%    \item{} SGI Altix  [?]
%  \end{itemize*}
%One of these actions will disable MPI
\begin{itemize*}
  \item{} Modify \iterm{topdir/CMakeLists.txt}:
\begin{code}
SET(QMC_MPI 0)
\end{code}
  \item{} Set \icode{QMC\_MPI} environment to 0, eg. \icode{export QMC\_MPI=0} for bash.
\end{itemize*}
\textbf{Note}: Shell environment variables take precedence over CMake settings.

\paragraph{Multithreading}
OpenMP is disabled by default in \iterm{CMakeLists.txt} but is automatically enabled if CMake detects that these compilers are being used:
\begin{itemize*}
  \item{} Intel compilers
  \item{} IBM VisualAge compilers
  \item{} GNU/OpenMP compilers $>$ 4.2.x on Linux 2.6.x kernels, Mac OS X 
\end{itemize*}
To enable OpenMP for other compilers, one of the following actions will be needed:
\begin{itemize*}
  \item{} Modify \iterm{topdir/CMakeLists.txt}:
\begin{code}
SET(QMC_OMP 1)
\end{code}
  \item{} Set \icode{QMC\_OMP} to 1, eg. \icode{export QMC\_OMP=1} for bash.
\end{itemize*}

[check this] If your machine has multiple cores, there is no need to disable OpenMP.  However, make sure to set the environment variables which control OpenMP runs.  Especially with MKL, set
\begin{code}
MKL_NUM_THREADS=1
MKL_SERIAL=YES
\end{code}
so that the blas/lapack calls DO NOT USE threaded version.

Note that the default number of threads on your machine may be set to the number of cores (or CPU units). It is always safe to set the number of threads yourself as, for example,
\begin{code}
export OMP_NUM_THREADS=1
\end{code}

%More on cmake
%cmake environment variables
In addition to \icode{QMC\_MPI} and \icode{QMC\_OMP}, there are few more environment variables used by QMCPACK and CMake to determine compiler-time options. \icode{QMC\_COMPLEX} (0 or 1) sets wavefunctions to take real or complex values.\footnote{One can use real wavefunction for the complex-enabled build but it will be extremely inefficient (4-8 times slower). However, complex wavefunctions cannot be used with real-enabled build.}  \icode{QMC\_BITS} (32 or 64) sets the OS bit size.  A change in any one of these variables will cause Make to rebuild everything instead of rebuilding just the modified source files.
%Each build should use identical variables. If the working shell has different variables from the previous build environments, cmake/make will rebuild everything which can take time. Out-of-source compilation becomes very useful to build different combinations.
%Note that separate executables have to be built for real and complex wavefunctions.  

%\subsection{Library dependencies}
%\subsection{Required utilities and libraries}
%In order to install QMCPACK, users have to install several required packages. These packages are included in standard Linux/cygwin distributions or can be downloaded by following the links. If these libraries are installed in standard directories, /usr /usr/local and /sw (Mac), no action is necessary. Alternatively, environment variables XYZ\_HOME should be set. Here, XYZ stands for the name of package; the build utility can locate the libraries and use them.

%With few exceptions, the build utility cmake will look for XYZ\_HOME/include for the header files and XYZ\_HOME/lib for the library files. When multiple environment variables apply to a library, e.g., blas/lapack, the library is searched according to the listed order. 
%intel mkl, gsl, boost, mpich/openmpi, openmp, cuda, etc.
%\begin{itemize}
%\item{} cmake   utility                 source and binary distribution on multiple platforms
%\item{} blas/lapack     Numerical library       MKL\_HOME, LAPACK, ATLAS         Alternatives: vendor-provided blas, e.g., ESSL
%\item{} hdf5    I/O     HDF5\_HOME, HDF\_HOME     source and binary distribution on multiple platforms
%\item{} libxml2         I/O     LIBXML2\_HOME    Standard library for Linux distributions
%\item{} boost   C++ standard libraries  BOOST\_HOME      Using only the header files. No need to compile anything. Simply download and unpack the package.
%\item{} einspline       c library for 3D bspline        see Using Einspline library     
%\item{} fftw    c library for FFT       FFTW\_HOME       http://www.fftw.org/
%
%\end{itemize}

\paragraph{Numerical Libraries}
%If lapack/blas or atlas is not in your standard path, do one of the following. ``location'' is where the libraries are located.
If LAPACK/BLAS or ATLAS is not in your library search path, set the environment variables \icode{LAPACK} to \icode{"-L/installdir -llapack -lblas"} and \icode{ATLAS} to\\ \icode{"-L/installdir -llapack -lf77blas -lcblas -latlas"}, changing \iterm{installdir} appropriately.

%For bash users,
%\begin{verbatim}
% export LAPACK="-L/location -llapack -lblas"
% export ATLAS="-L/location -llapack -lf77blas -lcblas -latlas" 
%\end{verbatim}
%For tcsh users,
%\begin{verbatim}
% setenv LAPACK "-L/location -llapack -lblas"
% setenv ATLAS "-L/location -llapack -lf77blas -lcblas -latlas"
%\end{verbatim}

\subsection{Configurations known to work}
\subsubsection{Intel 64 Abe cluster @ NCSA}
Abe uses SoftEnv to activate the configurations for the (otherwise dormant) software installed in it.  Create the file .soft in the home directory, containing the following lines:
\begin{term}
@default
@boost-1.42.0
@teragrid-basic
+intel-mkl
+libxml2
+hdf5-1.8.4
\end{term}
Set the following environment variables:
\begin{term}
CXX=mpicxx
CC=icc
LIBXML2\_HOME=/usr/local/libxml2-2.6.29
QMC\_BITS=64
QMC\_OMP=1
\end{term}

\subsubsection{Intel Mac OS X}
[I can try this on my laptop, with MacPorts]

Basic steps

    * Install cmake and hdf5 via fink (building from source is perfectly acceptable)
          o Set HDF5\_HOME as

\begin{verbatim}
            export HDF5_HOME=/sw
\end{verbatim}

    * libxml2 is most likely in the standard paths, i.e., /usr/include and /usr/lib
          o Otherwise, use fink and set

\begin{verbatim}
            export LIBXML2_HOME=/sw
\end{verbatim}

    * Download and unpack boost
          o Set BOOST\_HOME to the directory name (full path) of the boost package 

[edit] Few facts

    * Do not install lapack/blas via fink or port
    * cmake will use Mac Framework and link the libraries automatically 

[edit] Extras

    * The performance of plane-wave depends on blas library. Framework or GotoBlas do not perform as well as MKL.
    * If Intel compilers and MKL are available, set the environments as 

\begin{verbatim}
export CXX=icpc
export CC=icc
export MKL_HOME=/Library/Frameworks/Intel_MKL.framework/Versions/Current
\end{verbatim}

    * Then, cmake/make
    * Check DYLD\_LIBRARY\_PATH. Add the lines below (32->64 for 64bit OS) in the shell 

\begin{verbatim}
if [ -z "${LD_LIBRARY_PATH}" ]
then
  LD_LIBRARY_PATH=$MKL_HOME/lib/32; export LD_LIBRARY_
PATH
else
  LD_LIBRARY_PATH=$MKL_HOME/lib/32:$LD_LIBRARY_PATH; export LD_LIBRARY_PATH
fi

if [ -z "${DYLD_LIBRARY_PATH}" ]
then
  DYLD_LIBRARY_PATH=$MKL_HOME/lib/32:; export DYLD_LIBRARY_PATH
else
  DYLD_LIBRARY_PATH=$MKL_HOME/lib/32:$DYLD_LIBRARY_PATH; export DYLD_LIBRARY_PATH
fi
\end{verbatim}

\subsection{Tested environments}
QMCPACK is developed for large-scale high-performance computing systems. We update the status of QMCPACK on the HPC systems the developers have access to and share experiences in dealing with some quirkiness of each system.

In general, the quickest way to build and use QMCPACK is to use a toolchain file for each system. They are available in config directory with the distribution. Current list includes
\begin{itemize*}
\item{} AbeMvapich2.20091104.cmake : abe@ncsa
\item{} AbeMvapich2.cmake
\item{} AbeOpenMPI.cmake
\item{} BGToolChain.cmake : BlueGeneP
\item{} BluePrintXL.cmake : BluePrint@ncsa, IBM P5
\item{} JaguarGNU.cmake : jaguar@ornl, Cray XT
\item{} KrakenGNU.cmake : kraken@nics, Cray XT
\item{} LinuxIntel.cmake : generic for LINUX using Intel compilers
\item{} LOP6\_xlC\_xlc.cmake : huygens@sara, IBM P6, running LINUX, IBM XL compilers
\item{} LOP6\_xlC\_gcc.cmake : huygens@sara, IBM P6, running LINUX, IBM XLC compiler and GNU C compiler
\end{itemize*}
Once a toolchain file (mychain.cmake)  is selected,
\begin{verbatim}
cd build
cmake -DCMAKE_TOOLCHAIN_FILE=../config/mychain.cmake ..
cmake -DCMAKE_TOOLCHAIN_FILE=../config/mychain.cmake ..
make
\end{verbatim}
Feel free to make a local copy in build directory and modify it for your environment. For each platform, only the paths to the external libraries have to be modified. Where the toolchain file is located is not important.

\subsubsection{Cygwin}
Update qmcpack using svn, especially src/CMakeLists.txt
 Summary

    * cygwin 1.5.25-7
    * GCC 3.4.4 

 Libraries

    * cmake : cygwin update
    * libxml2 : cygwin update
          o Set LIBXML2\_HOME=/usr/include/libxml2 
    * boost : select boost-devel using cygwin update
          o Set BOOST\_HOME=/usr/include/boost-1\_33\_1
          o Or search the directory where boost/config.hpp resides 
    * HDF5
          o download source from HDF5 home page:
            wget ftp://ftp.hdfgroup.org/HDF5/current16/src/hdf5-1.6.7.tar.gz
                + Do not use version 1.8.x. 
          o unpack:
            tar zxf hdf5-1.6.7.tar.gz
          o configure and make:
            cd hdf5-1.6.7; ./configure --prefix=/usr/local; make; make install
          o Set HDF5\_HOME=/usr/local 

\subsection{Known version conflicts}
\section{A walk-through example with Si}
I guess Jaron knows more about this...
\subsection{Input file layout}
mention just the standard-ness of xml and postpone further discussion to next chapter
\subsection{Preparing wavefunctions}
\subsubsection{Wavefunction optimization}
\subsection{Your first VMC calculation}
\subsection{Your first DMC calculation}
\subsection{Understanding the output}
mention data analysis tool and postpone discussion for two chapters
%Obtaining QMCPACK
%  SVN instructions
%Building QMCPACK
%  Simple build instructions
%  Building with toolchain files (machine specific)
%  Building QMCPACK from scratch
%    Library Dependencies and where to obtain them
%    Detailed build instructions
%  Using the benchmark database to test your build
%  Known Good Configurations (combinations of library versions that work)
%  Known version conflicts
%  Who to contact if you are experiencing difficulties
