\chapter{Additional Tools}\label{chap:additional_tools}
QMCPACK provides a set of lightweight executables that address certain
common problems in QMC workflow and analysis.  These range from conversion utilities between 
different file formats and QMCPACK (e.g., \ishell{ppconvert} and \ishell{convert4qmc}),  
(qmc-extract-eshdf-kvectors) to postprocessing utilities (\ishell{trace-density} and \ishell{qmcfinitesize}) to many others.  In this chapter, we cover the use cases, syntax, and features of all additional tools provided with QMCPACK.  

\section{Initialization Tools}
  \subsection{qmc-get-supercell}

\section{Postprocessing}
  \subsection{qmca}
    \ishell{qmca} is a versatile tool to analyze and plot the raw data from QMCPACK \ishell{*.scalar.dat} files.
    It is a Python executable and part of the Nexus suite of tools.  It can be found in 
    \ishell{qmcpack/nexus/executables}. For details, see Section~\ref{sec:qmca}.
  \subsection{qmc-fit}
    \ishell{qmc-fit} is a curve fitting tool used to obtain statistical error bars on fitted parameters.
    It is useful for DMC time step extrapolation.  For details, see Section~\ref{sec:qmcfit}.
  \subsection{qdens}
    \ishell{qdens} is a command line tool to produce density files from QMCPACK's \ishell{stat.h5} output files.  For details, see Section~\ref{sec:qdens}.
  \subsection{qmcfinitesize}
    \ishell{qmcfinitesize} is a utility to compute many-body finite-size corrections to the energy.  It
    is a C++ executable that is built alongside the QMCPACK executable.  It can be found in 
    \ishell{build/bin}.

\section{Converters} 
\input{convert4qmc}    
\subsection{pw2qmcpack.x}\label{sec:pw2qmcpack}
\ishell{pw2qmcpack.x} is an executable that converts PWSCF wavefunctions to QMCPACK readable 
HDF5 format.  This utility is built alongside the QE postprocessing utilities.
This utility is written in Fortran90 and is distributed as a patch of the QE 
source code.  The patch, as well as automated QE download and patch scripts, can be found in 
\ishell{qmcpack/external_codes/quantum_espresso}.

pw2qmcpack can be used in serial in small systems and should be used in parallel with large systems for best performance. The K\_POINT gamma optimization is not supported.

\begin{lstlisting}[style=ESPRESSO,caption={Sample \ishell{pw2qmcpack.x} input file \ishell{p2q.in}}]
&inputpp
  prefix     = 'bulk_silicon'
  outdir     = './'
  write_psir = .false.
/
\end{lstlisting}

This example will cause \ishell{pw2qmcpack.x} to convert wavefunctions saved from PWSCF with the prefix ``bulk\_silicon.'' Perform the conversion via, for example:

\begin{shade}
mpirun -np 1 pw2qmcpack.x < p2q.in>& p2q.out
\end{shade}

Because of the large plane-wave energy cutoffs in the pw.x calculation required by accurate PPs and the large system sizes of interest, one limitation of QE can be easily reached:
that \ishell{wf_collect=.true.} results in problems of writing and loading correct plane-wave coefficients on disks by pw.x because of the 32 bit integer limits. Thus, \ishell{pw2qmcpack.x} fails to convert the orbitals for QMCPACK. Since the release of QE v5.3.0, the converter has been fully parallelized to overcome this limitation completely.

By setting \ishell{wf_collect=.false.} (by default \ishell{.false.} in v6.1 and before and \ishell{.true.} since v6.2), pw.x does not collect the whole wavefunction into individual files for each k-point but instead writes one smaller file for each processor.
By running \ishell{pw2qmcpack.x} in the same parallel setup (MPI tasks and k-pools) as the last scf/nscf calculation with pw.x,
the orbitals distributed among processors will first be aggregated by the converter into individual temporal HDF5 files for each k-pool and then merged into the final file.
In large calculations, users should benefit from a significant reduction of time in writing the wavefunction by pw.x thanks to avoiding the wavefunction collection.

pw2qmcpack has been included in the test suite of QMCPACK (see instructions about how to activate the tests in Section~\ref{sec:buildqe}).
There are tests labeled ``no-collect'' running the pw.x with the setting \ishell{wf_collect=.false.}
The input files are stored at \ishell{examples/solids/dft-inputs-polarized-no-collect}.
The scf, nscf, and pw2qmcpack runs are performed on 16, 12, and 12 MPI tasks with 16, 2, and 2 k-pools respectively.
\subsection{convertpw4qmc}
Convertpw4qmc is an executable that reads xml from a plane wave based DFT code and produces a QMCPACK readable 
HDF5 format wavefunction.  For the moment, this supports both QBox and Quantum Epresso

In order to save the wavefunction from QBox so that convertpw4qmc can work on it, one needs to add a line to the
QBox input like
\begin{shade}
save -text -serial basename.sample
\end{shade}
after the end of a converged dft calculation.  This will write an ascii wavefunction file and will avoid
QBox's optimized parallel IO (which is not currently supported).

After the wavefunction file is written (basename.sample in this case) one can use convertpw4qmc as follows:
\begin{shade}
convertpw4qmc basename.sample -o qmcpackWavefunction.h5
\end{shade}

This reads the Qbox wavefunction and performs the Fourier transform before saving to a QMCPACK eshdf format wavefunction.  Currently multiple k-points are supported, but due to difficulties with the qbox wavefunction file format, the single particle orbitals do not have their proper energies associated with them.  This means that when tiling from a primitive cell to a supercell, the lowest n single particle orbitals from all necessary k-points will be used.  This can be problematic in the case of a metal and this feature should be used with EXTREME caution.

In the case of quantum espresso, QE must be compiled with HDF support.  If this is the case, then an eshdf file can be generated by targeting the data-file-schema.xml file
generated in the output of quantum espresso.  For example, if one is running a calculation with outdir = 'out' and prefix='Pt' then the converter can be invoked as:
\begin{shade}
  convertpw4qmc out/Pt.save/data-file-schema.xml -o qmcpackWavefunction.h5
\end{shade}
Note, that this method is insensitive to parallelization options given to quantum espresso.  Additionally, it supports noncollinear magnetism and can be used to generate
wavefunctions suitable for qmcpack calculations with spin-orbit coupling.

\subsection{ppconvert}\label{sec:ppconvert}
\texttt{ppconvert} is a utility to convert PPs between different commonly used formats.
It is a stand-alone C++ executable that is not built by default but that is accessible via adding
\ishell{-DBUILD_PPCONVERT=1} to CMake and then typing \ishell{make ppconvert}.
Currently it converts CASINO, FHI, UPF (generated by OPIUM), BFD, and GAMESS formats to several other formats
including XML (QMCPACK) and UPF (QE). See all the formats via \ishell{ppconvert -h}.
For output formats requiring Kleinman-Bylander projectors, the atom will be solved with DFT
if the projectors are not provided in the input formats.
This requires providing reference states and sometimes needs extra tuning for heavy elements.
To avoid ghost states, the local channel can be changed via the \ishell{--local_channel} option. Ghost state considerations are similar to those of DFT calculations but could be worse if ghost states were not considered during the original PP construction.
To make the self-consistent calculation converge, the density mixing parameter may need to be reduced
via the \ishell{--density_mix} option.
Note that the reference state should include only the valence electrons.
One reference state should be included for each channel in the PP.
For example, for a sodium atom with a neon core, the reference state would be ``1s(1).''
\ishell{--s_ref} needs to include a 1s state, \ishell{--p_ref} needs to include a 2p state,
\ishell{--d_ref} needs to include a 3d state, etc. If not specified, a corresponding state with zero occupation is added.
If the reference state is chosen as the neon core, setting empty reference states ``'' is technically correct.
In practice, reasonable reference states should be picked with care.
For PP with semi-core electrons in the valence, the reference state can be long.
For example, Ti PP has 12 valence electrons. When using the neutral atom state,
\ishell{--s_ref}, \ishell{--p_ref}, and \ishell{--d_ref} are all set as ``1s(2)2p(6)2s(2)3d(2).''
When using an ionized state, the three reference states are all set as ``1s(2)2p(6)2s(2)'' or ``1s(2)2p(6)2s(2)3d(0).''
Unfortunately, if the generated UPF file is used in QE, the calculation may be incorrect because of the presence of ``ghost'' states. Potentially these can be removed by adjusting the local channel (e.g., by setting \ishell{--local_channel 1}, which chooses the p channel as the local channel instead of d.
For this reason, validation of UPF PPs is always required from the third row and is strongly encouraged in general. For example, check that the expected ionization potential and electron affinities are obtained for the atom and that dimer properties are consistent with those obtained by a quantum chemistry code or a plane-wave code that does not use the Kleinman-Bylander projectors.
\input{pseudopotentials}

\subsection{wftester}\label{sec:wftester}
While not really a stand-alone application, wftester (short for ``Wave Function Tester") is a helpful tool for testing pre-existing and experimental estimators and observables.  It provides the user with derived quantities from the Hamiltonian and wave function, but evaluated at a small set of configurations. 

The wftester is implemented as a QMCDriver, so one invokes \qmcpack in the normal manner with a correct input XML, the difference being the addition of an additional qmc input block.  This is the main advantage of this tool--it allows testing of realistic systems and realistic combinations of observables.  It can also be invoked before launching into optimization, VMC, or DMC runs, as it is a valid <qmc> block.

 As an example, the following code generates a random walker configuration and compares the trial wave function ratio computed in two different ways:

\begin{lstlisting}[style=QMCPXML,caption=The following executes the wavefunction ratio test in ``wftester" ]
  <qmc method="wftester">
    <parameter name="ratio">    yes    </parameter>
  </qmc>
\end{lstlisting}

Here's a summary of some of the tests provided:
\begin{itemize}
\item Ratio Test.  Invoked with \begin{verbatim}<parameter name="ratio">yes</parameter>\end{verbatim}  This computes the implemented wave function ratio associated with a single-particle move using two different methods.
\item Clone Test. Invoked with \begin{verbatim}<parameter name="clone">yes</parameter>\end{verbatim}  This checks the cloning of TrialWaveFunction, ParticleSet, Hamiltonian, and Walkers.  
\item Elocal Test.  Invoked with \begin{verbatim}<parameter name="printEloc">yes</parameter>\end{verbatim}  For an input electron configuration (can be random), print the value of TrialWaveFunction, LocalEnergy, and all local observables for this configuration.  
\item Derivative Test.  Invoked with \begin{verbatim}<parameter name="ratio">deriv</parameter>}\end{verbatim}  Computes electron gradients, laplacians, and wave function parameter derivatives using implemented calls and compares them to finite-difference results.   
\item Ion Gradient Test.  Invoked with \begin{verbatim}<parameter name="source">ion0</parameter>\end{verbatim}  Calls the implemented evaluateGradSource functions and compares them against finite-difference results.  
\item ``Basic Test".  Invoked with \begin{verbatim}<parameter name="basic">yes</parameter>\end{verbatim}  Performs ratio, gradient, and laplacian tests against finite-difference and direct computation of wave function values.  

\end{itemize}

The output of the various tests will be to standard out or ``wftest.000" after successful execution of qmcpack.  
 
