\chapter{Derivation of twist averaging efficiency}
\label{sec:app_ta_efficiency}
In this appendix we derive the relative statistical efficiency of 
twist averaging with an irreducible (weighted) set of k-points 
versus using uniform weights over an unreduced set of k-points 
(\emph{e.g.} a full Monkhorst-Pack mesh).

Consider the weighted average of a set of statistical variables 
$\{x_m\}$ with weights $\{w_m\}$:
\begin{align}
  x_{TA} = \frac{\sum_mw_mx_m}{\sum_mw_m}
\end{align} 
If produced by a finite quantum Monte Carlo run at a set of 
twist angles/k-points $\{k_m\}$, each variable mean $\mean{x_m}$ 
has a statistical error bar $\sigma_m$ and we can also obtain 
the statistical error bar of the mean of the twist averaged 
quantity $\mean{x_{TA}}$:
\begin{align}
  \sigma_{TA} = \frac{\left(\sum_mw_m^2\sigma_m^2\right)^{1/2}}{\sum_mw_m}
\end{align}
The error bar of each individual twist $\sigma_m$, is related to the 
autocorrelation time $\kappa_m$,  intrinsic variance $v_m$, and number 
of post-equilibration Monte Carlo steps $N_{step}$ in the following way:
\begin{align}
  \sigma_m^2=\frac{\kappa_mv_m}{N_{step}}
\end{align}
In the setting of twist averaging, the autocorrelation time and 
variance for different twist angles are often very similar across 
twists and we have
\begin{align}
  \sigma_m^2=\sigma^2=\frac{\kappa v}{N_{step}}
\end{align} 
If we define the total weight as $W$, \emph{i.e.} $W\equiv\sum_{m=1}^Mw_m$, 
for the weighted case with $M$ irreducible twists the error bar is
\begin{align}
  \sigma_{TA}^{weighted}=\frac{\left(\sum_{m=1}^Mw_m^2\right)^{1/2}}{W}\sigma
\end{align}
For uniform weighting with $w_m=1$ the number of twists is $W$ and 
we have
\begin{align}
  \sigma_{TA}^{uniform}=\frac{1}{\sqrt{W}}\sigma
\end{align}
We are interested in comparing the efficiency of choosing weights 
uniformly or based on the irreducible multiplicity of each twist angle 
for a given target error bar $\sigma_{target}$.  The number of Monte Carlo  
steps required to reach this target for uniform weighting is
\begin{align}
  N_{step}^{uniform} = \frac{1}{W}\frac{\kappa v}{\sigma_{target}^2}
\end{align}
while for non-uniform weighting we have
\begin{align}\label{eq:weighted_step}
  N_{step}^{weighted} &= \frac{\sum_{m=1}^Mw_m^2}{W^2}\frac{\kappa v}{\sigma_{target}^2} \nonumber\\
                  &=\frac{\sum_{m=1}^Mw_m^2}{W}N_{step}^{uniform}
\end{align}
The Monte Carlo efficiency is defined as 
\begin{align}
  \xi = \frac{1}{\sigma^2t}
\end{align}
where $\sigma$ is the error bar and $t$ is the total cpu time required 
for the Monte Carlo run.  

The main advantage made possible by irreducible twist weighting is to 
reduce the equilibration time overhead by having fewer twists, and 
hence fewer Monte Carlo runs to equilibrate.  In the context of twist 
averaging, the total cpu time for a run can be considered to be
\begin{align}
  t=N_{twist}(N_{eq}+N_{step})t_{step}
\end{align}
where $N_{twist}$ is the number of twists, $N_{eq}$ is the number of Monte 
Carlo steps required to reach equilibrium, $N_{step}$ is the number 
of Monte Carlo steps included in the statisical averaging as before, 
and $t_{step}$ is the wall clock time required to complete a single 
Monte Carlo step. For uniform weighting $N_{twist}=W$ while for irreducible 
weighting $N_{twist}=M$.

We can now calculate the relative efficiency ($\eta$) of irreducible vs. 
uniform twist weighting with the aim of obtaining a target error bar 
$\sigma_{target}$:
\begin{align}
  \eta &= \frac{\xi_{TA}^{weighted}}{\xi_{TA}^{uniform}} \nonumber \\
       &= \frac{\sigma_{target}^2t_{TA}^{uniform}}{\sigma_{target}^2t_{TA}^{weighted}} \nonumber \\
       &= \frac{W(N_{eq}+N_{step}^{uniform})}{M(N_{eq}+N_{step}^{weighted})} \nonumber \\
       &= \frac{W(N_{eq}+N_{step}^{uniform})}{M(N_{eq}+\frac{\sum_{m=1}^Mw_m^2}{W}N_{step}^{uniform})} \nonumber \\
       &= \frac{W}{M}\frac{1+f}{1+\frac{\sum_{m=1}^Mw_m^2}{W}f}
\end{align}
In this last expression, $f$ is the ratio of the number of usable 
Monte Carlo steps to the number that must be discarded during equilibration 
($f=N_{step}^{uniform}/N_{eq}$) and as before $W=\sum_mw_m$, which is the number of 
twist angles in the uniform weighting case.  It is important to recall 
that $N_{step}^{uniform}$ in $f$ is defined relative to uniform weighting; it is 
the number of Monte Carlo steps required to reach a target accuracy in the 
case of uniform twist weights.

The formula for $\eta$ above can be easily changed with the help of 
Eq. \ref{eq:weighted_step} to reflect the 
number of Monte Carlo steps obtained in an irreducibly weighted run 
instead.  A good exercise is to consider runs one has already completed 
with either uniform or irreducible weighting and calculate the 
expected efficiency change had the opposite type of weighting been used.

The break even point $(\eta=1)$ can be found at a usable step fraction of 
\begin{align}
  f=\frac{W-M}{M\frac{\sum_{m=1}^Mw_m^2}{W}-W}
\end{align}

The relative efficiency $(\eta)$ above is useful to consider in view of certain 
scenarios.  An important case is where the number of required sampling 
steps is no larger than the number of equilibration steps, \emph{i.e.} 
$f\approx 1$.  For a very simple case with 8 uniform twists with 
irreducible multiplicities of $w_m\in\{1,3,3,1\}$ ($W=8$, $M=4$), the 
relative efficiency of irreducible vs. uniform weighting is 
$\eta=\frac{8}{4}\frac{2}{1+20/8}\approx 1.14$.  In this case, 
irreducible weighting is about $14$\% more efficient than uniform weighting.

Another interesting case is where the number of sampling steps you can 
reach with uniform twists before wall clock time runs out is small 
relative to the number of equilibration steps ($f\rightarrow 0$). 
In this limit $\eta\approx W/M$.  For our 8 uniform twist example, this would 
result in a relative efficiency of $\eta=8/4=2$ making irreducible 
weighting twice as efficient.

A final case of interest is where the equilibration time is short 
relative to the available sampling time $(f\rightarrow\infty)$, 
giving $\eta\approx W^2/(M\sum_{m=1}^Mw_m^2)$.  Again for our simple example 
we find $\eta=8^2/(4\times 20)\approx 0.8$, with uniform weighting being 
$25$\% more efficient than irreducible.  

For this simple example, the crossover point for irreducible weighting being 
more efficient than uniform weighting is $f<2$, \emph{i.e.} when the 
available sampling period is less than twice the length of the equilibration 
period.  The expected efficiency ratio and crossover point should be checked 
for the particular case under consideration to inform the choice between   
twist averaging methods.



