\section{Jastrow Factors}
\label{sec:jastrow}

Jastrow factors are among the simplest and most effective ways of including
dynamical correlation in the trial many body wavefunction.  The resulting many body
wavefunction is expressed as the product of an antisymmetric (in the case
of Fermions) or symmetric (for Bosons) part and a correlating Jastrow factor
like so:
\begin{equation}
\Psi(\vec{R}) = \mathcal{A}(\vec{R}) \exp\left[J(\vec{R})\right]
\end{equation}

In this section we will detail the types and forms of Jastrow factor used 
in QMCPACK.  Note that each type of Jastrow factor needs to be specified using
its own individual \texttt{jastrow} XML element.  For this reason, we have repeated the
specification of the \texttt{jastrow} tag in each section, with specialization for the
options available for that given type of Jastrow.

\subsection{One-body Jastrow functions}
\label{sec:onebodyjastrow}
The one-body Jastrow factor is a form that allows for the direct inclusion
of correlations between particles that are included in the wavefunction with
particles that are not explicitly part of it.  The most common example of
this are correlations between electrons and ions.  

The Jastrow function is specified within a \texttt{wavefunction} element
and must contain one or more \texttt{correlation} elements specifying
additional parameters as well as the actual coefficients. Section
\ref{sec:1bjsplineexamples} gives examples of the typical nesting of
\texttt{jastrow}, \texttt{correlation}, and \texttt{coefficient} elements.

\subsubsection{Input Specification}

\begin{table}[h]
\begin{center}
\begin{tabular}{l c c c l }
\hline
\multicolumn{5}{l}{Jastrow element} \\
\hline
\bfseries name & \bfseries datatype & \bfseries values & \bfseries defaults  & \bfseries description \\
\hline
name & text &    & (required) & Unique name for this Jastrow function \\
type & text & One-body & (required) & Define a one-body function \\ 
function & text & Bspline & (required) & BSpline Jastrow \\
             & text & pade2 & & Pade form \\
             & text & \ldots & & \ldots \\
source & text & name & (required) & Name of attribute of classical particle set \\ 
print & text & yes / no & yes & Jastrow factor printed in external file?\\
  \hline
\multicolumn{5}{l}{elements}\\ \hline
& Correlation & & & \\ \hline
\multicolumn{5}{l}{Contents}\\ \hline
& (None)  & & &  \\ \hline
\end{tabular}
%\end{tabular*}
\end{center}
\end{table}

To be more concrete, the one-body Jastrow factors used to describe correlations
between electrons and ions take the form below
\begin{equation}
J1=\sum_I^{ion0}\sum_i^e u_{ab}(|r_i-R_I|)
\end{equation}
where I runs over all of the ions in the calculation, i runs over the electrons
and $u_{ab}$ describes the functional form of the correlation between them.
Many different forms of $u_{ab}$ are implemented in QMCPACK.  We will detail 
two of the most common ones below.
\include{jastrow_one_body_spline}
\include{jastrow_one_body_pade}
\include{jastrow_one_body_srcusp}

\subsection{Two-body Jastrow functions}
The two-body Jastrow factor is a form that allows for the explicit inclusion
of dynamic correlation between two particles included in the wavefunction.  It
is almost always given in a spin dependent form so as to satisfy the Kato cusp
condition between electrons of different spins\cite{kato}.

 The two body Jastrow function is specified within a \texttt{wavefunction} element
and must contain one or more correlation elements specifying additional parameters
as well as the actual coefficients.  Section~\ref{sec:2bjsplineexamples} gives 
examples of the typical nesting of \texttt{jastrow}, \texttt{correlation} and
\texttt{coefficient} elements.

\subsubsection{Input Specification}

\begin{table}[h]
\begin{center}
\begin{tabular}{l c c c l }
\hline
\multicolumn{5}{l}{Jastrow element} \\
\hline
\bfseries name & \bfseries datatype & \bfseries values & \bfseries defaults  & \bfseries description \\
\hline
name & text &    & (required) & Unique name for this Jastrow function \\
type & text & Two-body & (required) & Define a one-body function \\ 
function & text & Bspline & (required) & BSpline Jastrow \\
print & text & yes / no & yes & Jastrow factor printed in external file?\\
  \hline
\multicolumn{5}{l}{elements}\\ \hline
& Correlation & & & \\ \hline
\multicolumn{5}{l}{Contents}\\ \hline
& (None)  & & &  \\ \hline
\end{tabular}
%\end{tabular*}
\end{center}
\end{table}

The two-body Jastrow factors used to describe correlations between electrons take the form
\begin{equation}
J2=\sum_i^{e}\sum_{j>i}^{e} u_{ab}(|r_i-r_j|)
\end{equation}

The most commonly used form of two body Jastrow factor supported by the code is a splined
Jastrow factor, with many similarities to the one body spline Jastrow.

\include{jastrow_two_body_spline}

\include{jastrow_user}

\subsection{Long-ranged Jastrow factors}
While short-ranged Jastrow factors capture the majority of the benefit 
for minimizing the total energy and the energy variance, long-ranged 
Jastrow factors are important to accurately reproduce the short-ranged 
(long wavelength) behavior of quantities such as the static structure 
factor, and are therefore essential for modern accurate finite size 
corrections in periodic systems.

Below two types of long-ranged Jastrow factors are described.  The 
first (the k-space Jastrow) is simply an expansion of the one and/or 
two body correlation functions in plane waves, with the coefficients 
comprising the optimizable parameters.  The second type have few 
variational parameters and use the optimized breakup method of Natoli 
and Ceperley\cite{Natoli1995} (the Yukawa and Gaskell RPA Jastrows).


\subsubsection{Long-ranged Jastrow: k-space Jastrow}
The k-space Jastrow introduces explicit long-ranged dependence commensurate with the periodic supercell.  This Jastrow is to be used in periodic boundary conditions only.  

The input for the k-space Jastrow fuses both one and two-body forms into a single element and so they are discussed together here.  The one- and two-body terms in the k-Space Jastrow have the form:
\begin{align}
  J_1 &= \sum_{G\ne 0}b_G\rho_G^I\rho_{-G} \\
  J_2 &= \sum_{G\ne 0}a_G\rho_G\rho_{-G}
\end{align}
Here $\rho_G$ is the Fourier transform of the instantaneous electron density:
\begin{align}
  \rho_G=\sum_{n\in electrons}e^{iG\cdot r_n}
\end{align}
and $\rho_G^I$ has the same form, but for the fixed ions. In both cases the coefficients are restricted to be real, though in general the coefficients for the one-body term need not be.  See section~\ref{sec:feature_kspace_jastrow} for more detail.

Input for the k-space Jastrow follows the familar nesting of \texttt{jastrow-correlation-coefficients} elements, with attributes unique to the k-space Jastrow at the \texttt{correlation} input level.

\FloatBarrier
\begin{table}[h]
\begin{center}
\begin{tabularx}{\textwidth}{l l l l l X }
\hline
\multicolumn{6}{l}{\texttt{jastrow type=kSpace} element} \\
\hline
\multicolumn{2}{l}{parent elements:} & \multicolumn{4}{l}{\texttt{wavefunction}}\\
\multicolumn{2}{l}{child  elements:} & \multicolumn{4}{l}{\texttt{correlation}}\\
\multicolumn{2}{l}{attributes}  & \multicolumn{4}{l}{}\\
   &   \bfseries name     & \bfseries datatype & \bfseries values          & \bfseries default  & \bfseries description \\
   & \texttt{type}$^r$    &  text              & \textbf{kSpace}           &                    & Must be kSpace           \\
   & \texttt{name}$^r$    &  text              & \textit{anything}         & 0                  & Unique name for Jastrow \\
   & \texttt{source}$^r$  &  text              & \texttt{particleset.name} &                    & Ion particleset name\\
  \hline
\end{tabularx}
\end{center}
\end{table}
\FloatBarrier

\FloatBarrier
\begin{table}[h]
\begin{center}
\begin{tabularx}{\textwidth}{l l l l l X }
\hline
\multicolumn{6}{l}{\texttt{correlation} element} \\
\hline
\multicolumn{2}{l}{parent elements:} & \multicolumn{4}{l}{\texttt{jastrow type=kSpace}}\\
\multicolumn{2}{l}{child  elements:} & \multicolumn{4}{l}{\texttt{coefficients}}\\
\multicolumn{2}{l}{attributes}  & \multicolumn{4}{l}{}\\
   &   \bfseries name           & \bfseries datatype & \bfseries values  & \bfseries default  & \bfseries description \\
   & \texttt{type}$^r$          &  text              & \textbf{One-Body},\textbf{Two-Body}    &                     & Must be One-Body/Two-Body     \\
   & \texttt{kc}$^r$            &  real              & kc$\ge$ 0                                & 0.0                 & k-space cutoff in a.u. \\
   & \texttt{symmetry}$^o$      &  text              & crystal,isotropic,none                 & crystal             & Symmetry of coefficients\\
   & \texttt{spinDependent}$^o$ &  boolean           & yes,no                                 & no                  & \textit{No current function} \\
  \hline
\end{tabularx}
\end{center}
\end{table}
\FloatBarrier

\FloatBarrier
\begin{table}[h]
\begin{center}
\begin{tabularx}{\textwidth}{l l l l l X }
\hline
\multicolumn{6}{l}{\texttt{coefficients} element} \\
\hline
\multicolumn{2}{l}{parent elements:} & \multicolumn{4}{l}{\texttt{correlation}}\\
\multicolumn{2}{l}{child  elements:} & \multicolumn{4}{l}{\textit{None}}\\
\multicolumn{2}{l}{attributes}  & \multicolumn{4}{l}{}\\
   &   \bfseries name     & \bfseries datatype & \bfseries values  & \bfseries default   & \bfseries description \\
   & \texttt{id}$^r$      &  text              & \textit{anything} &     cG1/cG2         & Label for coeffs     \\
   & \texttt{type}$^r$    &  text              & \texttt{Array}    &   0                 & Must be Array \\
\multicolumn{2}{l}{body text}  & \multicolumn{4}{l}{}\\
   &                           & \multicolumn{4}{l}{The body text is a list of real values for the parameters.}     \\
  \hline
\end{tabularx}
\end{center}
\end{table}
\FloatBarrier


Additional information:
\begin{itemize}
  \item{It is normal to provide no coefficients as an initial guess.  The number of coefficients will be automatically calculated according to the k-space cutoff + symmetry and set to zero. }
  \item{Providing an incorrect number of parameters also results in all parameters being set to zero.}
  \item{There is currently no way to turn optimization on/off for the k-space Jastrow.  The coefficients are always optimized.}
  \item{Spin dependence is currently not implemented for this Jastrow.}
  \item{\texttt{kc}: Parameters with G vectors magnitudes less than \texttt{kc} are included in the Jastrow.  If \texttt{kc} is zero, it is the same as excluding the k-space term.}
  \item{\texttt{symmetry=crystal}: Impose crystal symmetry on coefficients according to the structure factor.}
  \item{\texttt{symmetry=isotropic}: Impose spherical symmetry on coefficients according to G-vector magnitude.}
  \item{\texttt{symmetry=none}: Impose no symmetry on the coefficients.}
\end{itemize}


\begin{lstlisting}[style=QMCPXML,caption=k-space Jastrow with one- and two-body terms.]
  <jastrow type="kSpace" name="Jk" source="ion0">
     <correlation kc="4.0" type="One-Body" symmetry="cystal">
        <coefficients id="cG1" type="Array">                  
        </coefficients>
     </correlation>
     <correlation kc="4.0" type="Two-Body" symmetry="crystal">
        <coefficients id="cG2" type="Array">                  
        </coefficients>
     </correlation>
  </jastrow>
\end{lstlisting}

\begin{lstlisting}[style=QMCPXML,caption=k-space Jastrow with one-body term only.]
  <jastrow type="kSpace" name="Jk" source="ion0">
     <correlation kc="4.0" type="One-Body" symmetry="cystal">
        <coefficients id="cG1" type="Array">                  
        </coefficients>
     </correlation>
  </jastrow>
\end{lstlisting}

\begin{lstlisting}[style=QMCPXML,caption=k-space Jastrow with two-body term only.]
  <jastrow type="kSpace" name="Jk" source="ion0">
     <correlation kc="4.0" type="Two-Body" symmetry="crystal">
        <coefficients id="cG2" type="Array">                  
        </coefficients>
     </correlation>
  </jastrow>
\end{lstlisting}




\subsubsection{Long-ranged Jastrows: Gaskell RPA and Yukawa forms}
\label{sec:twobodyjastrowlr}
\textbf{NOTE: The Yukawa and RPA Jastrows do not work at present 
and are currently being revived.  Please contact the developers if 
you are interested in using them.} 

The exact Jastrow correlation functions contain terms which have a 
form similar to the Coulomb pair potential.  In periodic systems 
the Coulomb potential is replaced by an Ewald summation of the 
bare potential over all periodic image cells.  This sum is often 
handled by the optimized breakup method\cite{Natoli1995} and this 
same approach is applied to the long-ranged Jastrow factors in QMCPACK.

There are two main long-ranged Jastrow factors of this type 
implemented in QMCPACK: the Gaskell RPA\cite{Gaskell1961,Gaskell1962} 
form and the Yukawa\cite{Ceperley1978} form.  Both of these forms 
were used by Ceperley in early studies of the electron gas\cite{Ceperley1978}, 
but they are also appropriate starting points for general solids. 

The Yukawa form is defined in real space.  It's long-range form is 
formally defined as
\begin{align}
  u_Y^{PBC}(r) = \sum_{L\ne 0}\sum_{i<j}u_Y(\abs{r_i-r_j+L})
\end{align}
with $u_Y(r)$ given by
\begin{align}
  u_Y(r) = \frac{a}{r}\left(1-e^{-r/b}\right)
\end{align}
In QMCPACK a slightly more restricted form is used:
\begin{align}
  u_Y(r) = \frac{r_s}{r}\left(1-e^{-r/\sqrt{r_s}}\right)
\end{align}
here ``$r_s$'' is understood to be a variational parameter.

The Gaskell RPA form--which contains correct short/long range limits 
and minimizes the total energy of the electron gas within the RPA--is 
defined directly in k-space:
\begin{align}
  u_{RPA}(k) = -\frac{1}{2S_0(k)}+\frac{1}{2}\left(\frac{1}{S_0(k)^2}+\frac{4m_ev_k}{\hbar^2k^2}\right)^{1/2}
\end{align}
where $v_k$ is the Fourier transform of the Coulomb potential and 
$S_0(k)$ is the static structure factor of the non-interacting 
electron gas:
\[
  S_0(k) = \left.
  \begin{cases}
    1 &  k>2k_F \\
    \frac{3k}{4k_F}-\frac{1}{2}\left(\frac{k}{2k_F}\right)^3 & k<2k_F
  \end{cases}
  \right.
\]
When written in atomic units, RPA Jastrow implemented in QMCPACK has the 
form
\begin{align}
  u_{RPA}(k) = \frac{1}{2N_e}\left(-\frac{1}{S_0(k)}+\left(\frac{1}{S_0(k)^2}+\frac{12}{r_s^3k^4}\right)^{1/2}\right)
\end{align}
Here ``$r_s$'' is again a variational parameter and $k_F\equiv(\tfrac{9\pi}{4r_s^3})^{1/3}$.

For both the Yukawa and Gaskell RPA Jastrows, the default value for $r_s$ is $r_s=(\tfrac{3\Omega}{4\pi N_e})^{1/3}$.


\FloatBarrier
\begin{table}[h]
\begin{center}
\begin{tabularx}{\textwidth}{l l l l l X }
\hline
\multicolumn{6}{l}{\texttt{jastrow type=Two-Body function=rpa/yukawa} element} \\
\hline
\multicolumn{2}{l}{parent elements:} & \multicolumn{4}{l}{\texttt{wavefunction}}\\
\multicolumn{2}{l}{child  elements:} & \multicolumn{4}{l}{\texttt{correlation}}\\
\multicolumn{2}{l}{attributes}  & \multicolumn{4}{l}{}\\
   &   \bfseries name        & \bfseries datatype & \bfseries values             & \bfseries default  & \bfseries description   \\
   & \texttt{type}$^r$       &  text              & \textbf{Two-Body}            &                    & Must be Two-Body   \\
   & \texttt{function}$^r$   &  text              & \textbf{rpa}/\textbf{yukawa} &                    & Must be rpa or yukawa   \\
   & \texttt{name}$^r$       &  text              & \textit{anything}            & RPA\_Jee            & Unique name for Jastrow \\
   & \texttt{longrange}$^o$  &  boolean           & yes/no                       & yes                & Use long-range part     \\
   & \texttt{shortrange}$^o$ &  boolean           & yes/no                       & yes                & Use short-range part    \\
\multicolumn{2}{l}{parameters}  & \multicolumn{4}{l}{}\\
   & \texttt{rs}$^o$         &  rs                & $r_s>0$                      & $\tfrac{3\Omega}{4\pi N_e}$ & Avg. elec-elec distance \\
   & \texttt{kc}$^o$         &  kc                & $k_c>0$                      & $2\left(\tfrac{9\pi}{4}\right)^{1/3}\tfrac{4\pi N_e}{3\Omega}$ & K-space cutoff\\
  \hline
\end{tabularx}
\end{center}
\end{table}
\FloatBarrier


\begin{lstlisting}[style=QMCPXML,caption=Two body RPA Jastrow with long- and short-ranged parts.]
<jastrow name=''Jee'' type=''Two-Body'' function=''rpa''>
</jastrow>
\end{lstlisting}



% J1 RPA (intended for electron-proton system)
%   source = particleset.name
%   function = RPA
%   name = anything [Jep]
%   rs = >0 [-1]
%   kc = >0 [-1]
% 
% J2 RPA
%   attributes
%     function = yukawa or rpa
%     name = anything [RPA_Jee]
%     longrange = yes/no [yes]
%     shortrange = yes/no [yes]
%   parameters
%     rs = >0 [-1]  3\Omega/4\pi N_e
%     kc = >0 [-1]  2 (9\pi/4)^1/3 * 4\pi N_e/3\Omega



\subsection{Three-body Jastrow functions}
Explicit three-body correlations can be included in the wavefunction via the three-body Jastrow factor.
The three-body electron-electron-ion correlation function ($u_{\sigma\sigma'I}$) currently used in \qmcpack is identical to the one proposed in \cite{Drummond2004}:
\begin{eqnarray}
u_{\sigma\sigma'I}(r_{\sigma I},r_{\sigma'I},r_{\sigma\sigma'}) &= \sum_{\ell=0}^{M_{eI}}\sum_{m=0}^{M_{eI}}\sum_{n=0}^{M_{ee}}\gamma_{\ell mn} r_{\sigma I}^\ell r_{\sigma'I}^m r_{\sigma\sigma'}^n \\
   &\times \left(r_{\sigma I}-\frac{r_c}{2}\right)^3 \Theta\left(r_{\sigma I}-\frac{r_c}{2}\right) \nonumber \\
   &\times \left(r_{\sigma' I}-\frac{r_c}{2}\right)^3 \Theta\left(r_{\sigma' I}-\frac{r_c}{2}\right) \nonumber
\end{eqnarray}
Here $M_{eI}$ and $M_{ee}$ are the maximum polynomial orders of the
electron-ion and electron-electron distances, respectively,
$\{\gamma_{\ell mn}\}$ are the optimizable parameters (modulo
constraints), $r_c$ is a cutoff radius, and $r_{ab}$ are the distances
between electrons or ions $a$ and $b$. i.e. The correlation function
is only a function of the interparticle distances and not a more
complex function of the particle positions, $\mathbf{r}$. As indicated by the
$\Theta$ functions, correlations are set to zero beyond a distance of
$r_c/2$ in either of the electron-ion distances and the largest
meaningful electron-electron distance is $r_c$.  This is the
highest-order Jastrow correlation function currently implemented.


Today, solid state applications of \qmcpack usually utilize one and
two-body B-spline Jastrow functions, with calculations on heavier
elements often also using the three-body term described above.

\paragraph{Example use case}
Here is an example of H2O molecule. After optimizing one and two body Jastrow factors, add the following block in the wavefunction.
The coefficients will be filled zero automatically if not given.
\begin{lstlisting}[style=QMCPXML]
<jastrow name="J3" type="eeI" function="polynomial" source="ion0" print="yes">
  <correlation ispecies="O" especies="u" isize="3" esize="3" rcut="10">
    <coefficients id="uuO" type="Array" optimize="yes"> </coefficients>
  </correlation>
  <correlation ispecies="O" especies1="u" especies2="d" isize="3" esize="3" rcut="10">
    <coefficients id="udO" type="Array" optimize="yes"> </coefficients>
  </correlation>
  <correlation ispecies="H" especies="u" isize="3" esize="3" rcut="10">
    <coefficients id="uuH" type="Array" optimize="yes"> </coefficients>
  </correlation>
  <correlation ispecies="H" especies1="u" especies2="d" isize="3" esize="3" rcut="10">
    <coefficients id="udH" type="Array" optimize="yes"> </coefficients>
  </correlation>
</jastrow>
\end{lstlisting}
