\section{Main input file}\label{input.intro}

A valid input file for QMCPACK should look like

\begin{lstlisting}[language=xml,escapeinside={<:}{:>},
emph={simulation},emphstyle=\bfseries\color{blue}]
  <?xml version="1.0"?>
  <simulation>
    <project id="mysimulation" series="0"/><:\coref{1}{project.qr}:>
    <random parallel="1" seed="-1"/><:\coref{2}{random.qr}:>
    <particleset/> <:\coref{3}{particleset.qr}:>
    <wavefunction/> <:\coref{4}{wavefunction.qr}:>
    <hamiltonian/> <:\coref{5}{hamiltonian.qr}:>
    <qmc/> <:\coref{6}{qmc.qr}:>
  </simulation> 
\end{lstlisting}

\textbf{The order of the xml elements in an input file is important, e.g., {\tt
<qmc>} cannot come before other elements.} The alllowed multiplicity of each
xml element is given in [ ]s. 

\begin{description}
%[leftmargin=1cm,style=sameline]
\item[{\hyperref[project.qr]{\conum{1}}}]
sets the title and sequence number of a simulation. [0..1]
\item[{\hyperref[random.qr]{\conum{2}}}]
assigns the seed of the random number generator (using Singleton pattern). [0..1] 
\item[{\hyperref[particleset.qr]{\conum{3}}}]
creates a set of Particles. [1..*]
\item[{\hyperref[wavefunction.qr]{\conum{4}}}]
creates a many-{}body wavefunction for a target particleset.  [1..*]
\item[{\hyperref[hamiltonian.qr]{\conum{5}}}]
creates a many-{}body Hamiltonian for a target particleset. [1..*]
\item[{\hyperref[qmc.qr]{\conum{6}}}]
executes a QMC algorithm defined by the {\tt method} attribute. [1..*]
\end{description}

When the document is not well formed, e.g., it has mismatched tags, the xml
parser will issue errors and stop execution of QMCPACK. A common tool like {\tt
xmllint} can be used to check if your input file is well formed.  

%This section covers the key input parameters to perform a QMC simulation. 
%At the minimum, QMC simulation requires i) a set of \emph{particles}, e.g.,
%electrons, whose configuration will be updated according to a QMC method, ii) a
%many-{}body wave function for the quantum particle set, and iii) many-{}body
%Hamiltonian to evaluate the local energy.i

\subsection{General simulation parameters}

QMCPACK and other related programs process the elements in this section to set
the simulation parameters, such as the root of the output files and random
number generators. 

%\subsubsection{project} \label{project.sec}
\begin{description}
\item[\textbf{project}{}]\label{project.sec}
In order to maintain multiple QMC runs for a system,
\hyperlink{project.element}{project} is used to specify the title of the
simulation and the sequence index. The \texttt{title} and \texttt{id} are used
for the root name of the output files. 
 
Each recording engines will add a postfix. The postfix typically tells you what
data the output contains and its format.  For example, a simulation with the
project as
\begin{lstlisting}[language=XML,emph={project},emphstyle=\bfseries\color{blue}]
  <project id="H2.STO.mo" series="0"/>
\end{lstlisting} 
will produce \nolinkurl{H2.STO.mos000.scalar.dat} and several file names. Any
file whose postfix is \nolinkurl{dat} can be plotted with any graphic package
that accepts multiple column data, e.g., gnuplot, xmgrace and dataspork. See
\hyperlink{output.sec}{output} for the standard output of QMCPACK.  Although it
is an optional element, specifying the \textbf{project} is strongly
recommended.

%\item[\textbf{application}{}]
%This \emph{optional} element \hyperlink{application.element}{application} is
%intended to record the version of the code, related publications and other
%information that can be used for data mining. 

\item[\textbf{random}{}]
This \emph{optional} element \hyperlink{random.element}{random} sets the
properties of the global random number generator using Singleton pattern. The
default setting is
  
\begin{lstlisting}[language=XML,emph={random},emphstyle=\bfseries\color{blue}]
  <random parallel="1" seed="-1"/>
\end{lstlisting}
 to assign independent random seeds based on the hardware clock of the
 execution. The state of the random number generator is stored in a hdf5 for
 checkpoint and is used for the subsequent run.  Specifically, we use
 \href{http://www.boost.org/boost/random/mersenne_twister.hpp}{
 mersenne-{}twister} random number generator.
\end{description}
   
\subsection{Definition of a QMC system}

%A QMC system is a collection of physical objects, such as Particle sets
%(\hyperref[particleset.sec]{particleset}), trial many-{}body wavefunctions
%(\hyperref[wavefunction.sec]{wavefunction}), and Hamiltonian operators
%(\hyperref[hamiltonian.sec]{hamiltonian}).

Before performing any QMC simulation, we need to define and create the physical
objects. At the minimum, QMCPACK requires a Quantum Particle Set (denoted as
\emph{target particleset}), a trial wavefunction for the \emph{target
particleset}, and a Hamiltonian operator which defines a local energy operator
for the \emph{target particleset}.  Since the most common application  of
QMCPACK deals with the electronic structure of solids and molecules,  we use
\emph{electrons} and \emph{target particleset} interchangeably to denote the
Quantum Particle Set for a QMCPACK run.
 
\begin{description}
\item[\textbf{particleset}{}]
defines a set of Particles which are grouped together.  For example, a
\hyperlink{particleset.element}{particleset} defines an \emph{N-{}}electron
system for the electronic structure calculations.
 
\begin{lstlisting}[language=XML,emph={particleset},emphstyle=\bfseries\color{blue}]
  <particleset name="e">
    <group name="u" size="3">
      <parameter name="charge">-1</parameter>
    </group>
    <group name="d" size="3">
      <parameter name="charge">-1</parameter>
    </group>
  </particleset> 
\end{lstlisting}
A molecular system or crystalline system in a unit cell will comprise another
\texttt{particleset}.  The attribute \texttt{name} is used to name each
\texttt{particleset}.
 
\item[\textbf{wavefunction}{}]\label{wavefuncion.intro}
defines a trial wavefunction for a \emph{target particleset}.  The \emph{target
particleset} should be defined before used by the wavefunction.
 
\begin{lstlisting}[language=XML,emph={wavefunction},emphstyle=\bfseries\color{blue}]
  <wavefunction name="psi0" target="e">
    <jastrow name="Jee" type="Two-Body" spin="yes" function="pade">
      <parameter id="juu_b" name="B">0.821683</parameter>
    </jastrow>
  </wavefunction> 
\end{lstlisting}

\item[\textbf{hamiltonian}{}]
defines a total Hamiltonian for a \emph{target particleset}. The \emph{target
particleset}  and \emph{wavefunction} should be defined before used by the
\texttt{hamiltonian}.
 
\begin{lstlisting}[language=XML,emph={hamiltonian},emphstyle=\bfseries\color{blue}]
  <hamiltonian name="h0" target="e" wavefunction="psi0">
    <pairpot name="ElecElec" type="coulomb" source="e"/>
    <pairpot name="Coulomb" type="coulomb" source="i"/>
    <pairpot name="IonIon" type="coulomb" source="i" target="i"/>
  </hamiltonian> 
\end{lstlisting}
\end{description}

\subsection{Running  QMC}

\begin{description}
\item[\textbf{qmc}{}]
defines the actions on a QMC system, i.e., performing one of several QMC algorithms
to calculate the properties of the \emph{target particleset}.
 
\begin{lstlisting}[language=XML,emph={qmc},emphstyle=\bfseries\color{blue}]
  <qmc method="dmc" move="pbyp" target="e">
    <parameter name="time_step">0.05</parameter>
  </qmc> \end{lstlisting}
\end{description}
